Social inequality in mortality among adults and elderly in northern Sweden 1851--2013\footnote{Paper presented at the session 'Health and Mortality in Arctic Populations'. The Ninth International Congress on Arctic Social Sciences, Umeå, June 8--12, 2017}

### 2017-05-31 icass.Rnw
Updating education and income

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha} 
\citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--64 and 65--89 years of age, the first age group representing people that
were still mainly in the workforce, and the second those that mainly were 
retired. The model controls for period, marital
status and whether the individual resides in the urban or rural part. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}.

We analyse both regions together but perform separate analyses for men and
women and for the age groups 40--64 and 65--89. The model in the analyses
include social class, sex and marital status. Each decade
during the 20th century gets separate analyses, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period.

@@

@@

The analysis for the period 1990--2005 for the whole population 40 years and older is based on separate analyses with income and education as explanatory variables and with controls of period 
and marital status. See Figure~\ref{fig:ploteduc}.

@@

Education had a strong, statistically significant, effect on survival, as expected.

@@

Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival. The differences must be considered large, and it follows almost a continuous improvement according to increasing levels.

\subsection{Cardiovascular death}

@@

Illustrating differences for cause-specific mortality is conveniently done by comparing 
cumulative hazards, see Figure~\ref{fig:ploteducheart}, where educational effects on 
cardiovascular mortality is illustrated. The pattern is exactly the same as for total 
mortality.

@@

\subsection{Cancer death}

@@

Although mortality in cancer is lower than mortality in cardiovascular diseases,
the pattern is the same: The more education, the less mortality. See 
Figure~\ref{fig:ploteduccancer}.

@@

Higher incomes result in lower mortality. This is also what could be expected, 
and conforms to what we know of social health differences in present-day Sweden. 
Notice however that we cannot establish a causal relationship between 
income and mortality. Bad health, for instance, may be a confounder.

\subsection{Cardiovascular death}

@@

The overall pattern is the same for cardiovascular mortality, high income pays off.
<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1990--2013.">>=
oldpar <- par(mfrow = c(1, 2))

@@

The same goes for cancer mortality, but the relative differences are smaller. See Figure~\ref{fig:plotinccancer}.

@@

One of the major findings in this study relate to gender. As several researchers have pointed out, demography has often ignored power relations and different roles within families \citep{watkins93, mack95, jans07}. By investigating possible differences in mortality patterns according to sex, we have identified gendered patterns within the family. Different patterns between the sexes may be caused by discrimination on the female part in the household. We know that the society was
strongly segregated according to gender and that women were discriminated
and had less space for action. In our case, we do not find this had negative consequences for women, indicating that the relation between sexessociety was characterised by distinct gendered expectations on behaviour.

>>>>

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha} 
\citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--64 and 65--89 years of age, the first age group representing people that
were still mainly in the workforce, and the second those that mainly were 
retired. The model controls for marital
status and whether the individual resides in the urban or rural part. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}. Each period are analysed separately, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period.

% We analyse both regions together but perform separate analyses for men and
% women and for the age groups 40--64 and 65--89. The model in the analyses
% include social class, sex and marital status.

@@

%\subsection{All causes of death}

@@

In this section we analyse differences according to education level for the period 1990--2013 for the whole population 40 years and older. Education is a good indicator of both human and cultural capital, but aer also strongly associated with income and social class. We present analyses on all cause mortality (survival curves) as well as on the cause of death groups of cardiovascular diseases and cancers (cumulative hazards). The pattern is clear in total mortality as well as cause specific mortality, and this is valid for both men and women, see Figures~\ref{fig:ploteduc},~\ref{fig:ploteducheart} and~\ref{fig:ploteduccancer}. Education had a strong, statistically significant, effect on survival, as expected. Those with the lowest education levels hade the highest mortality, while those with highest level of education had by far the best survival. The differences must be considered large, and it follows a continuous improvement according to increasing levels.

@@

@@

%\subsection{Cardiovascular death}

@@

% Illustrating differences for cause-specific mortality is conveniently done by comparing 
% cumulative hazards, see Figure~\ref{fig:ploteducheart}, where educational effects on 
% cardiovascular mortality is illustrated. The pattern is exactly the same as for total 
% mortality.

@@

%\subsection{Cancer death}

@@

% Although mortality in cancer is lower than mortality in cardiovascular diseases,
% the pattern is the same: The more education, the less mortality. See 
% Figure~\ref{fig:ploteduccancer}.

@@

Finally, we present analyses of different income levels for the age group 40 and above and the period 1986--2005. Disposable individual income must be considered as a robust measure of economic condition, and increases the possibility to allocate these economic resources towards health promoting living conditions and behaviours. We categorize the income according to quartiles. As expected, we find that higher incomes result in lower mortality, both when it comes to total mortality as to deaths from cardiovascular diseases, see Figures~\ref{fig:plotincome} and~\ref{fig:plotincheart}. This is also what could be expected, and conforms to what we know of social health differences in present-day Sweden. We also find indications of a similar difference for cancers but less clear, particularly for male cancer mortality, see Figure~\ref{fig:plotinccancer}. Notice however that we cannot establish a causal relationship between income and mortality. Bad health may be a confounder, causing both low incomes and higher death risks.

%\subsection{Cardiovascular death}

@@

%The overall pattern is the same for cardiovascular mortality, high income pays off.
<<plotincheart, fig.cap="Cumulative hazards by income and sex, cardiovascular death, 1990--2013.">>=
oldpar <- par(mfrow = c(1, 2))

@@

% The same goes for cancer mortality, but the relative differences are smaller. See Figure~\ref{fig:plotinccancer}.

@@

One of the major findings in this study relate to gender. As several researchers have pointed out, demography has often ignored power relations and different roles within families \citep{watkins93, mack95, jans07}. By investigating possible differences in mortality patterns according to sex, we have identified gendered patterns within the family. Different patterns between the sexes may be caused by discrimination on the female part in the household. We know that the society was
strongly segregated according to gender and that women were discriminated
and had less space for action. In our case, we do not find this to have negative consequences for women, indicating that the relation between sexes in the society was characterised by distinct gendered expectations on behaviour.
